From big data to deep insight in developmental science

### 2015-09-02 wiley-big-data-devel/big-data.tex
final draft

\documentclass[letterpaper,man,apacite]{apa6}

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[hyphens]{url}
\usepackage{hyperref}

\hypersetup{breaklinks=true}

@@

%\abstract{}

\authornote{Rick O. Gilmore is in the Department of Psychology, The Pennsylvania State University, University Park, PA 16802, rogilmore@psu.edu.
Any opinions, findings, and conclusions or recommendations expressed in the material contributed here are those of the author and do not necessarily reflect the views of the National Science Foundation, the Eunice Kennedy Shriver National Institute of Child Health and Human Development, or the Society for Research in Child Development.}

@@

A search on the term ``big data'' yields more than 49 million hits on Google, more than 100,000 results on Google Scholar, and 13 million hits on Bing.
The results return in less than a second.
A search of the term using Google's Ngram viewer (\url{https://books.google.com/ngrams}) that indexes terms in digitized books shows the first appearance of the term right before 1900 with a steady rise in frequency from the 1950s to just after 2000.
So, whether measured by search engine matches of electronic documents available on the Internet, scholarly documents, or digitized books, the term ``big data'' gets widespread use now.
That use has grown substantially over the past several decades.
The fact that these basic facts about a particular phrase can be determined in an instant speaks to the rapid pace of change in networked computing, search engines, and databases.
Most of the tools that enable it have been created in the last 20 years.
``Big data'' has become a significant cultural phenomenon \cite{borgman_big_2015,boyd_critical_2012}, with frequent feature articles in the popular \cite{lohr_big_2012,Marcus2013} and specialist press \cite{HBR2015,Press2013a}.

In this review, I focus on how the increased availability of and interest in analyzing large, complex, multi-level, multi-measure data sets has influenced, continues to influence, and promises to alter the study of human development.
I begin by asking what makes data especially big, and what implications the size, density, or complexity of datasets might have for understanding human development.
Then, I survey a set of case studies that illustrate how large, complex data sets have been and are being used currently to reveal the complexities of developmental processes.
I close by discussing a range of issues that the emergence of ``big data'' approaches to developmental science pose for the future of the field.

I argue against the claim that the emergence of an era of ``big data'' signals the end of theory \cite{anderson_end_2008} while acknowledging the lure of that notion to some.
Nor are ``big data'' analyses all that new in developmental psychology.
Rather, while some the promises of ``big data'' to revolutionize scientific understanding are exaggerated \cite{boyd_critical_2012}, developmental science tackles questions that can benefit substantially from larger, richer, more widely shared, and more readily inter-operable datasets.
Thus, whether significant novel insights into human development emerge from the era of ``big data'' depend not just on the size or complexity of the datasets, but on how widely and openly the data are shared, and how readily able researchers are able to combine or link datasets across levels of analysis.
In turn, these specific innovations depend largely on small, probably manageable, but nonetheless thorny problems related to policy, scientific culture, individual researcher behavior, publisher priorities, and research funding levels.
Technology may accelerate the ``big data'' era, but the challenges it poses may turn out to be the least important where advancing research in developmental psychology is concerned.

\subsection{What Does `Big Data` Mean in Developmental Science?}

According to Laney \citeyear{laney01controlling3v} the volume, velocity, and variety of data streams make data big.
Of course, general statements about the total quantity of data generated per day \cite{ibm_2015} make little sense outside of a specific research context.
What is high volume data to a a developmental psychologist, an archive of 10 terabytes (TB) of video and flat-file data, for example, is tiny to a physicist working on the Large Hadron Collider (LHC) that generates 30 petabytes in a year (\url{http://home.web.cern.ch/about/computing}).
Similarly, the measure of volume matters.
The Inter-university Consortium for Political and Social Research (ICPSR) (\url{https://www.icpsr.umich.edu}), one of the largest data repositories for data from the social sciences, consists of more than 500,000 files in 16 specialized data collections.
Yet, until the recent acquisition of video data from the Gates Foundation-funded Methods of Effective Teaching (MET) Project (\url{http://www.metproject.org}), the large number of files in the repository totaled about 10 TB of digital storage.

Even more important than the quantity of information stored is the kind of information: What are the targets of inquiry?
The range of phenomena covered by developmental science spans divergent levels of analysis in space, from genes to geography, and time, from microseconds to millenia, all aimed at answering two questions: What develops? Why does it develop this way?
In seeking answers, developmental scientists have long recognized the importance of multiple, nested influences on developmental processes arising at different temporal and spatial scales \cite{elman_rethinking_1998,gottlieb_normally_1998,oyama_ontogeny_2000,vygotsky_mind_1980}.
For social and behavioral scientists, the goal is to describe change in the \emph{behavior} or \emph{beliefs} of individuals or groups of individuals and families.
Data about neighborhoods, schools, and broader social, cultural, political and environmental influences primarily inform thinking about the development of individual or family behavior.
Similarly, the assessment of biological data -- genes, hormone levels, physiological responses, brain activity, body dimensions, brain structure, disease or disorder status -- inform thinking about how these within-person factors influence changes within individuals' behavior across time.

Consequently, the aspects of data `volume' pertinent to most developmental scientists include the number of participants or families or the number measurement time points.
Some notable datasets have hundreds or thousands of participants even though samples of tens of participants are more common.
High volume datasets enable more precise estimation of small effects, especially for qualities or conditions rare among individuals.
Similarly, the aspects of data `velocity' most relevant to developmental researchers are those related to the frequency or spacing of measurements.
Velocity can span many orders of magnitude, from physiological measurements collected at millisecond time scales to longitudinal research spanning years or decades.
High volume or velocity data may inform the estimation of trends within or between people across time \cite{rietveld_replicability_2014}.
`Variety' encompasses the range of measurement types employed across developmental research: biological, behavioral, contextual, and cultural/historical/evolutionary and the use of multiple measurements to address the same underlying construct.
To the ``three Vs'', variability, or the rate of change over time, and complexity, the mutual interdependence of individual measures, might also be added.

The collection, management, and analysis of data high in volume, velocity, variety, variability, and complexity poses significant practical challenges for data collection, capture, storage, sharing, analysis, transfer, visualization, and analysis.
Big data may also magnify the challenges facing researchers in maintaining participant privacy, in part because the more data that are collected, the more likely it is that individual identities can be discovered \cite{sweeney_identifiability}.
Big data also pose theoretical challenges -- e.g, how do micro-scale factors influence macro-scale phenomena?
Nevertheless, big data offer the possibility of tackling some of the most profound and vexing questions in the field if the relationships among data components can be revealed in ways that do not undermine ethical research principles.
Greater openness and more widespread sharing of research data than currently exists will also be essential to realize this scientific promise.

\section{Big datasets in developmental research}

Existing big datasets in developmental science fall into one of three broad categories depending on who collected the data and what mandates or restrictions apply to who may access it.
Many of the largest existing datasets available to developmental researchers come from data collected and disseminated by government entities like the U.S. Census Bureau and the U.S. Department of Education.
The collection and dissemination of these datasets is mandated by law, subject to provisions of a complex set of statutory and regulatory requirements designed to protect individual respondents' identities.
Government-collected datasets tend to be the most open and widely available to researchers, but by their very design and legal mandate the datasets are not intended for answering questions about individuals, families, or small groups.
Datasets collected by individual researchers or teams of researchers employed by institutions, usually colleges or universities, who are funded, usually by government or private funders, to collect large datasets form the next group.
These datasets tend to be somewhat smaller in size, but significantly more varied in terms of the types of information collected, the duration or intensity of data collection, the means of collection, the extent to which data are made available to individuals outside the research team, and the impact on scholarship.
The extent to which datasets collected by government-grant-funded research teams are shared with other researchers depends on a set of factors including institutional research ethics and data privacy policies, funder and journal requirements, and individual researcher's support for data sharing.
Finally, there are large scale datasets collected by private entities for business purposes that include sharing data about individual subscribers or users.
The data collected by private entities about individuals largely concerns consumer behavior, as the data customers are other commercial entities seeking to sell products or services to individual users.
But, commercial entities are beginning to collect and store health and fitness related activity (e.g., Fitbit, etc.) 
These datasets are sometimes made available to academic researchers, but the policies that govern data access are under the control of the entities that provide the services.

@@

Some of the largest and most widely used datasets in developmental science are population-based, often with a focus on health.
The National Health and Nutrition Examination Survey (NHANES; \url{http://www.cdc.gov/nchs/nhanes.htm}) began collecting data in 1959 with a focus on characterizing the health and nutritional status of U.S. infants, children, and adults. 
The annual survey combines interviews about demographic, socioeconomic, dietary and health related topics and physical examinations involving medical, dental, and physiological measurements, as well as laboratory tests.
The study is cross-sectional and assesses a nationally representative sample of about 5,000 persons located across the country, with 15 counties surveyed annually.
Many will be familiar with some of NHANES's products such as standardized child growth charts, and the closely associated National Youth Fitness Survey (NYFS; \url{http://www.cdc.gov/nchs/nnyfs.htm}).
NYFS collected data on physical activity and fitness levels in U.S. children ages 3 to 15 in 2012 in order to provide an evaluation their overall fitness.
Much of the NHANES and NYFS data is available to the public through the National Center for Health Statistics at the U.S. Centers for Disease Control.
However, access to some data are restricted, available only to individuals via one of the 19 Federal Statistical Research Data Centers (RDC) locations across the U.S.
A Google Scholar search in mid-2015 shows more than 110,000 publications mention the NHANES dataset.

The National Longitudinal Survey of Youth (NLSY; \url{http://www.bls.gov/nls/}) provides an example of a large-scale study that has examined the same cohort -- almost 10,000 American youth born between 1957-64 -- on a annual basis since the study's beginning in 1979.
Hosted by the U.S. National Bureau of Labor Statistics (BLS), NLSY data are now available for up to 25 survey rounds on questions relating to employment and related issues.
The NLSY has also been extended to encompass women born in 1922-37 and 1943-55 and men born 1906-21 and 1941-52.
Most NLSY and NLS data are available free of charge to by means of a web-based data portal to which users must apply for access.
Access to geographic-related variables requires special permission.
The Center for Human Resource Research (CHRR) at The Ohio State University and the National Opinion Research Center (NORC) at the University of Chicago administer both the data collection and repository functions.
A Google Scholar search in mid-2015 shows that more than 20,000 publications cite the NLSY.

The Early Childhood Longitudinal Study (ECLS; \url{https://nces.ed.gov/ecls/}) is another large-scale population-based study that has followed the same groups of young people over time focusing on child development, school readiness, and early school experiences. 
The ECLS consists of three cohorts, one focused on a sample of children measured beginning at birth and two cohorts following children from kindergarten.
The birth cohort a sample of children born in 2001 and followed from birth through kindergarten entry, and unusual among comparable studies includes data from fathers. 
One of the kindergarten cohorts studied approximately 22,000 children who entered school in 1998-99 and were followe through the eighth grade. 
A second cohort entered kindergarten in 2010-11 and is being followed through the fifth grade.
The project has received funding from the National Center for Educational Statistics (https://nces.ed.gov/ecls/), and the data are archived at ICPSR.
ECLS consists of both publicly available and restricted data.
A Google Scholar search generates more than 8,000 hits, of which the ICPSR archive can verify 167 project-related publications.

There are several large government-collected and shared datasets with information about employment patterns and educational attainment in children and youth.
The Bureau of Labor Statistics reports data about youth employment and unemployment patterns derived from the Census Bureau's Current Population Survey (\url{http://www.bls.gov/cps/demographics.htm}).
The U.S. Department of Education's National Center for Educational Statistics (NCES) collects and analyzes statistics on American primary, secondary, and post-secondary education, and many of the datasets are either public; data with individually identifiable data are available to researchers under a restricted use agreement.
The Trends in International Math and Science Study (TIMSS) has studied mathematics and science achievement in 4th and 8th grade for more than 20 years (\url{http://timss.bc.edu/}).
TIMSS is funded by the U.S. Department of Education, with domestic and international operations coordinated by a consortium of academic, association, and commercial partners.
The 2001 TIMSS assessment involved more than 20,000 U.S. students from more than 1,000 schools plus data from more than 60 countries.
The next TIMSS data collection is in 2015 and will include students in grades 4, 8 and 12.
Its sister study, the Progress in International Reading Literacy Study (PIRLS) has been studying reading achievement at the 4th grade.
Both studies provide data available for public download.

In addition to TIMSS, there are several examples of large-scale longitudinal studies focused on children's growth, health, and education from outside the U.S.
The World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS; \url{http://www.who.int/childgrowth/en/}) collected growth and related data from 8,500 children from widely different ethnic backgrounds and cultural settings, specifically Brazil, Ghana, India, Norway, Oman, and the U.S.
MGRS data provide body mass index, height for children 5-19, and weight data norms for children 5-10, in addition to other anthropomorphic measurements.
Launched in 2011, the Japan Environment and Children's Study (JECS; \url{http://www.env.go.jp/en/chemi/hs/jecs/}) is a birth cohort study involving 100,000 parent-child pairs with the goal of evaluating the impact of various environmental factors on children's health and development. 
Pilot studies started in 2008 and the hundred thousandth participating mother was enrolled in March 2014. 
Health outcomes and exposure measurements will continue until the participating children become 13 years old.
The study has produced a handful of publications, and no data sharing plans have been announced.
Several large-scale birth cohort studies have been undertaken in the U.K., including the National Survey of Health \& Development and the National Child Development Study (NCDS; \url{http://www.cls.ioe.ac.uk}).
NCDS is an ongoing longitudinal study that seeks to follow the lives of all those living in Great Britain who were born in one particular week in 1958.
Originally focused on improving infant mortality, the study has expanded to focus on factors predicting wellbeing across the lifespan.
Measures include genetic, health, education, social, and economic factors.
The study is funded by the British government and data are available by registration from the U.K. Data Service (\url{http://ukdataservice.ac.uk/get-data/key-data/cohort-and-longitudinal-studies}).
The British Cohort Studies have generated more than 3,500 publications to-date.
Finally, the Organization for Economic Cooperation and Development (OECD) publishes the PISA comparative educational dataset (\url{http://pisa2000.acer.edu.au/index.php}), which generates more than 5,000 search hits.

Not all large-scale birth cohort studies succeed, of course.
A notable failure in the U.S. is the National Children's Study (\url{http://www.nichd.nih.gov/research/NCS/Pages/default.aspx}).
Authorized by the Children's Health Act of 2000, the NCS would have followed 100,000 children and their parents from before birth to age 21. 
However, the NIH Director decided to close the NCS on December 12, 2014 following the recommendations of an advisory panel. 
Questionnaire, physical measures, biospecimens and environmental sample data from up to 5,726 participants in the NCS Vanguard study were collected in 2009-2014 prior to the NCS study's closure.
Those data are slated for release in a data archive sometime in 2015.

While no means exhaustive in its coverage, several themes emerge from this sample of U.S. and international population-based studies.
Datasets are large in terms of the volume of participants sampled, ranging in the thousands to tens of thousands.
Some studies involve tens of sample occasions per individual over extended periods of time, and show significant variety in the measures collected.
Most data are available to the public via web-based download or browsing/analysis portal, although access to data deemed sensitive may require a specific application and approval and possibly travel to a Research Data Center where special security provisions apply.
Finally, the datasets have generated research publications that range from the hundreds to the tens of thousands, suggesting that the collection, curation, and preservation of these resources has had a significant impact on scientific discovery.
We turn next to datasets whose collection and management is under the control of individual, usually academic investigators.

@@

Several big data studies in developmental science involve partnerships between government agencies and academic investigators.
Others involve large-scale data collections that are initiated and managed by individual investigators.
These projects are diverse in focus and methodology, including population-based studies with a health focus, to behavior genetics, brain imaging, cognition and temperament, and education and employment.

@@

A population-based example is the Panel Study on Income Dynamics (PSID; \url{https://psidonline.isr.umich.edu}) is large-scale population based survey study focusing on employment, income, wealth, expenditures, health, marriage, childbearing, child development, philanthropy, and education.
PSID began in 1968 with a U.S. nationally representative sample of 5,000 families and more than 18,000 individuals.
As of mid-2015, PSID had generated more than 3,900 publications. 
Much of data is available publicly to registered users and agree to the terms of use conditions that focus on ensuring that participants are not reidentified and that users properly cite and acknowledge use of the data. 
In addition, some restricted data elements -- e.g., geospatial data, school-level identifiers -- are only available subject to a formal restricted use agreement between the investigator and the data owner, the University of Michigan.

The NICHD Study of Early Child Care and Youth Development (SECCYD; \url{https://www.nichd.nih.gov/research/supported/Pages/seccyd.aspx}) is another population-based public/private partnership.
SECCYD prospectively followed the experiences of 1,364 infants born at 10 locations throughout the United States in 1991.
The goal was to learn more about the kind and quality of child care experiences and the effects of different child care settings on developmental outcomes. 
Measures included video, surveys and interviews, behavioral tests, biomarkers such as salivary cortisol, blood pressure, anthropometric and demographic information.
The study included multiple respondents, including the child, parents, caregivers, clinicians, and child care center directors.
Unlike the Ad Health study, data in SECCYD access are restricted, available only by application, with approval granted on a case-by-case basis.
Staff at the National Institute of Child Health and Human Development manage the data and access approval process.
The SECCYD has generated more than 214 publications as of mid-2015.

The National Longitudinal Study of Adolescent Health (Ad Health; \url{http://www.nichd.nih.gov/news/releases/Pages/adolescent.aspx}) represents a third example.
Ad Health began as a survey study undertaken in response to a Congressional mandate designed to measure the multiplicity of influences on adolescents' health and unhealthy behaviors. 
The study was carried out in two phases, the first involving surveys of 90,000 7-12th graders at 145 schools around the U.S. during the 1994-1995 school year.
The second phase involved 20,000 participants, where both the target adolescents and their parents underwent interviews at two time points separated by a year. 
The Ad Health cohort has subsequently been followed into young adulthood with four in-home interviews, the most recent in 2008, when the sample was aged 24-32.
During Wave IV, biomarker data from a blood test has made available data about glucose, hemoglobin (HbA1c) levels, blood lipids, and related measures. 
The Ad Health data are archived at ICSPR and are available for public use.
As of mid-2015, more than 5,000 publications had arisen from the Add Health dataset.

Investigator-initiated and managed, the Family Life Project (FLP; \url{http://flp.fpg.unc.edu}) took a population-based approach to studying the early development of children living in rural, largely poor communities in North Carolina and Pennsylvania. 
A birth cohort of 800 children were studied over the first 3 years of life in a series of home, child care visits and phone calls when the infants were 2, 6, 15, 24, and 36 months of age. 
Data collected included interviews, questionnaires, videotaped interactions, and biological samples yielding information about economic and community factors, family, home, language and cognition, stress, and genotype.
Data from Phase I of the FLP are archived at ICPSR, but are subject to restrictions on access to the sensitive and identifiable nature of the data.
The FLP site lists about 50 publications through 2013.

>>>>

\documentclass[letterpaper,man,apacite,natbib]{apa6}

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[hyphens]{url}
%\usepackage{hyperref}

% \hypersetup{breaklinks=true}

@@

\abstract{The use of the term ``big data'' has grown substantially over the past several decades, and is now widespread. 
In this review, I ask what makes data ``big'' and the implications of size, density, or complexity of datasets for the science of human development.
A survey of existing data sets illustrates how large, complex, multi-level, multi-measure data have been used to reveal the complexities of human development.
``Big data'' poses significant technical, policy, ethics, transparency, cultural, and conceptual issues that must be addressed.
Most big developmental science data are currently hard to find and cumbersome to access; the field lacks a culture of data sharing; and there is no consensus about who owns or should control research data.
But, these barriers are rapidly dissolving.
Developmental researchers are finding new ways to collect, manage, store, share, and enable others to reuse data about the multiple facets of human development.
This promises a future in which ``big data'' can lead to deep insight about some of the most vexing questions in behavioral science.
}

\authornote{Rick O. Gilmore is in the Department of Psychology, The Pennsylvania State University, University Park, PA 16802, rogilmore@psu.edu.
This work was supported by the National Science Foundation (BCS-1238599), the Eunice Kennedy Shriver National Institute of Child Health and Human Development (U01-HD-076595), and the Society for Research in Child Development.
Any opinions, findings, and conclusions or recommendations expressed in the material contributed here are those of the author and do not necessarily reflect the views of the National Science Foundation, the Eunice Kennedy Shriver National Institute of Child Health and Human Development, or the Society for Research in Child Development.
I thank Karen E. Adolph for her comments on the manuscript.}

@@

A search on the term ``big data'' yields more than 49 million hits on Google, more than 100,000 results on Google Scholar, and 13 million hits on Bing.
The results return in less than a second.
A search of the term using Google's Ngram viewer (\url{https://books.google.com/ngrams}) that indexes terms in digitized books shows the first appearance of the term around 1900 with a steady rise in frequency from the 1950s to around 2000.
Clearly, as measured by search engine matches of electronic documents on the Internet, scholarly documents, or digitized books, the use of the term ``big data'' has grown substantially over the past several decades, and is now widespread.
Moreover, the fact that these basic statistics about a particular phrase can be determined in an instant speaks to the rapid progress in networked computing, search engines, and databases.
Most of the tools that enable it have been created in the last 20 years.
``Big data'' has become a significant cultural phenomenon \cite{borgman_big_2015,boyd_critical_2012}, with frequent feature articles in the popular \cite{lohr_big_2012,Marcus2013} and specialist press \cite{HBR2015,Press2013a}.

In this review, I show how increased availability of and interest in big data sets promises to alter the study of human development.
I begin by asking what makes data ``big'' and the implications of size, density, or complexity of datasets for understanding human development.
Then, I survey case studies to illustrate how large, complex, multi-level, multi-measure data sets have been used to reveal the complexities of human development.
I conclude by discussing key questions that ``big data'' approaches pose for the future of developmental science pose.

I disagree with the claim that an era of ``big data'' signals the end of theory \cite{anderson_end_2008}.
Nor are ``big data'' analyses especially new in developmental science.
Although some claims about ``big data'' revolutionizing scientific understanding are exaggerated \cite{boyd_critical_2012}, developmental science tackles questions that can benefit substantially from larger, richer, more widely shared, and more readily inter-operable datasets.
Significant novel insights into human development emerging from the era of ``big data'' depend on the size, density, and complexity of the datasets, on how widely and openly the data are shared, and on how readily researchers are able to combine or link datasets across levels of analysis.
In turn, these specific innovations depend largely on small, probably manageable, but nonetheless thorny problems related to policy, scientific culture, individual researcher behavior, publisher priorities, and research funding levels.
Technology may accelerate the ``big data'' era, but the challenges it poses turn out to be relatively unimportant for advancing research in developmental psychology.

\subsection{What Does ``Big Data'' Mean in Developmental Science?}

According to Laney \citeyear{laney01controlling3v} the volume, velocity, and variety of data streams make data big.
Of course, general statements about the total quantity of data generated per day \cite{ibm_2015} make little sense outside of a specific research context.
High volume data for a developmental psychologist--an archive of 10 terabytes (TB) of video and flat-file data, for example--is tiny to a physicist working on the Large Hadron Collider (LHC) that generates 30 petabytes in a year (\url{http://home.web.cern.ch/about/computing}).
Similarly, what is ``big'' depends on how volume is measured.
The Inter-university Consortium for Political and Social Research (ICPSR) (\url{https://www.icpsr.umich.edu}), one of the largest data repositories for data from the social sciences, consists of more than 500,000 files in 16 specialized data collections.
Yet, until the recent acquisition of video data from the Gates Foundation-funded Methods of Effective Teaching (MET) Project (\url{http://www.metproject.org}), the entire repository totaled about 10 TB of digital storage.

Even more important than the quantity of information stored is the kind of information: What are the targets of inquiry?
The range of phenomena covered by developmental science spans multiple levels of analysis in space (from genes to geography) and time (from microseconds to millennia) all aimed at answering two questions: ``What develops?'' and  ``Why does it develop this way?''
In seeking answers, developmental scientists have long recognized the importance of multiple, nested influences on developmental processes arising at different temporal and spatial scales \cite{elman_rethinking_1998,gottlieb_normally_1998,oyama_ontogeny_2000,vygotsky_mind_1980}.
For social and behavioral scientists, the goal is to describe change in the psychological processes of individuals, groups of individuals, or families.
Data about neighborhoods, schools, and broader social, cultural, political and environmental influences primarily inform thinking about the development of individual or family behavior.
Similarly, the assessment of biological data--genes, hormones, physiological responses, brain activity, body dimensions, brain structure, disease or disorder status--inform thinking about how within-person factors influence changes behavior.

Consequently, the aspects of data \emph{volume} pertinent to most developmental scientists include the number of participants or families and/or the number of measurement time points.
Some datasets have hundreds or thousands of participants.
Such high volume datasets enable more precise estimation of small effects, especially for rare qualities or conditions.
Similarly, the aspects of data \emph{velocity} most relevant to developmental researchers relate to the frequency or spacing of measurements.
Velocity can span many orders of magnitude, from physiological measurements collected at millisecond time scales to longitudinal research spanning years or decades.
High volume or velocity data may inform the estimation of trends within and between people across time \cite{rietveld_replicability_2014}.
\emph{Variety} encompasses the range of measurement types employed across developmental research--Biological, behavioral, contextual, and cultural/historical/evolutionary--and the use of different types of measurements to address the same underlying construct.
In addition to the ``three Vs'', intra-individual \emph{variability} and complexity, or the mutual interdependence of individual measures, might also be included.

The collection, management, and analysis of data high in volume, velocity, variety, variability, and complexity poses significant practical challenges for data collection, capture, storage, sharing, analysis, transfer, visualization, and analysis.
Big data may also magnify the challenges facing researchers in maintaining participant privacy, in part because the more data that are collected, the more likely it is that individual identities can be discovered \cite{sweeney_identifiability}.
Big data also pose theoretical challenges: For example, how do micro-scale factors influence macro-scale phenomena?
Nevertheless, big data offer the possibility of tackling some of the most profound and vexing questions in the field if the relations among data components can be revealed in ways that do not undermine ethical research principles.
Greater openness and more widespread sharing of research data than is current practice will be essential to realize this scientific promise.
Developmental researchers can look to the recent past for examples of big datasets that have already been collected and in some cases, widely shared for inspiration.

\section{Big Datasets in Developmental Research}

Existing big datasets in developmental science fall into one of three broad categories depending on who collected the data and what mandates or restrictions apply to who may access it.
Many of the largest existing datasets available to developmental researchers come from data collected and disseminated by government entities such as the U.S. Census Bureau and the U.S. Department of Education.
The collection and dissemination of these datasets is mandated by law and subject to provisions of a complex set of statutory and regulatory requirements designed to protect individual respondents' identities.
Government-collected datasets tend to be the most open and widely available to researchers, but by their very design and legal mandate the datasets are not intended for answering questions about individuals, families, or small groups.
Datasets collected by individual researchers or teams of researchers employed by institutions, usually colleges or universities, who are funded, usually by government or private funders, to collect large datasets form the next group.
These datasets tend to be somewhat smaller in size, but significantly more varied in terms of the types of information collected, the duration or intensity of data collection, the means of collection, the extent to which data are made available to individuals outside the research team, and the impact on scholarship.
The extent to which datasets collected by government-grant-funded research teams are shared with other researchers depends on a set of factors including institutional research ethics and data privacy policies, funder and journal requirements, and individual researcher's support for data sharing.
Finally, some large-scale datasets are collected by private entities for business purposes and may include sharing data about individual subscribers or users.
The data collected by private entities about individuals largely concern consumer behavior because the data customers are other commercial entities seeking to sell products or services to individual users.
But, commercial entities are beginning to collect and store health and fitness related activity (e.g., Fitbit, etc.) 
These datasets are sometimes made available to academic researchers, but the policies that govern data access are under the control of the commercial entities that provide the services.

@@

Some of the largest and most widely used datasets in developmental science are population-based, often with a focus on health.
The National Health and Nutrition Examination Survey (NHANES; \url{http://www.cdc.gov/nchs/nhanes.htm}) began collecting data in 1959 on the health and nutritional status of U.S. infants, children, and adults. 
The annual survey combines interviews with a geographically diverse representative sample of about 5,000 people.
NHANES focuses on demographic, socioeconomic, dietary and health-related topics and physical examinations involving medical, dental, and physiological measurements, and laboratory tests of behavior.
The products include standardized child growth charts and the closely associated National Youth Fitness Survey (NYFS; \url{http://www.cdc.gov/nchs/nnyfs.htm}), which reports physical activity and fitness levels in 3 to 15-year-olds
Much of the NHANES and NYFS data are available to the public through the National Center for Health Statistics at the U.S. Centers for Disease Control.
However, access to some data is restricted, available only to individuals via one of the 19 Federal Statistical Research Data Centers locations across the U.S.
By mid-2015 more than 110,000 publications had cited the NHANES dataset.

In contrast to the cross-sectional NHANES, The National Longitudinal Survey of Youth (NLSY; \url{http://www.bls.gov/nls/}) has examined the same cohort--almost 10,000 American youth born between 1957 and 1964--on a annual basis since the study began in 1979.
NLSY data are now available for 25 survey rounds on questions relating to employment and other issues; extensions encompass wider age ranges.
Most NLSY and NLS data are available free of charge by means of a web-based data portal to which users must apply for access, but access to geographic-related variables requires special permission.
By mid-2015 more than 20,000 publications had cited the NLSY.

The Early Childhood Longitudinal Study (ECLS; \url{https://nces.ed.gov/ecls/}) is another longitudinal population-based study 
It focuses on child development, school readiness, and early school experiences in separate cohorts of children followed either from birth or from kindergarten that in some cases exceed 20,000 individuals.
ECLS consists of both publicly available and restricted data archived at ICPSR.
A Google Scholar search generates more than 8,000 hits, of which the ICPSR archive can verify 167 project-related publications.

Several large government-collected and shared datasets contain information about employment patterns and educational attainment in children and youth.
The Bureau of Labor Statistics reports data about youth employment and unemployment patterns derived from the Census Bureau's Current Population Survey (\url{http://www.bls.gov/cps/demographics.htm}).
The U.S. Department of Education's National Center for Educational Statistics (NCES) collects and analyzes statistics on U.S. primary, secondary, and post-secondary education, and many of the datasets are either public; data with personally identifiable data are available to researchers under a restricted use agreement.
The Trends in International Math and Science Study (TIMSS) has studied mathematics and science achievement in 4th and 8th grade for more than 20 years (\url{http://timss.bc.edu/}), with more than 20,000 U.S. students included in the 2001 sample.
De-identified data from TIMSS and its sister study, the Progress in International Reading Literacy Study (PIRLS; \url{http://timssandpirls.bc.edu}) are available for public download.

Some large-scale longitudinal studies focused on children's growth, health, and education include participants from outside the U.S.
The World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS; \url{http://www.who.int/childgrowth/en/}) collected physical growth and related data from 8,500 children in Brazil, Ghana, India, Norway, Oman, and the U.S.
The Japan Environment and Children's Study (JECS; \url{http://www.env.go.jp/en/chemi/hs/jecs/}) involves 100,000 parent-child pairs with the goal of evaluating the impact of environmental factors on children's health and development.
Large-scale birth cohort studies in the U.K. include the National Survey of Health \& Development and the National Child Development Study (NCDS; \url{http://www.cls.ioe.ac.uk}).
NCDS is an ongoing longitudinal study that follows everyone in Great Britain who was born in one particular week in 1958, focusing on factors predicting wellbeing across the lifespan.
NCDS data are available by registration from the U.K. Data Service (\url{http://ukdataservice.ac.uk/get-data/key-data/cohort-and-longitudinal-studies}).
The British Cohort Studies have generated more than 3,500 publications to-date.
The Organization for Economic Cooperation and Development (OECD) publishes the PISA comparative educational dataset (\url{http://pisa2000.acer.edu.au/index.php}), which generates more than 5,000 search hits.

Not all large-scale birth cohort studies succeed.
A notable failure is the National Children's Study (\url{http://www.nichd.nih.gov/research/NCS/Pages/default.aspx}).
Authorized by the Children's Health Act of 2000, the NCS would have followed 100,000 children prenatally until age 21.
However, the NIH Director decided to close the NCS in 2014 following the recommendations of an advisory panel. 
Questionnaire, physical measures, biospecimens and environmental sample data from up to 5,726 participants in the NCS Vanguard study were collected in 2009-2014 prior to study closure.
Those data are slated for release in a data archive sometime in 2015.

Several themes emerge from this sample of U.S. and international population-based studies.
Datasets are large in terms of the volume of participants sampled, ranging in the thousands to tens of thousands.
Some studies involve dozens of samples per individual over extended periods of time, and show significant variety in the measures collected.
Most data are available to the public via web-based download or browsing/analysis portal, although access to data deemed sensitive may require a specific application and approval and possibly travel to a Research Data Center where special security provisions apply.
Finally, the datasets have generated research publications that range from the hundreds to the tens of thousands, suggesting that the collection, curation, and preservation of these resources has had a significant impact scientific discovery.
I turn next to datasets whose collection and management is under the control of individual investigators.

@@

Several big data studies in developmental science involve partnerships between government agencies and academic investigators.
Others involve large-scale data collections initiated and managed by individual investigators and funded by government or foundation sources.
These projects are diverse in focus and methods, including population-based studies with a health focus, behavior genetics, brain imaging, cognition and temperament, and education and employment.

@@

The Panel Study on Income Dynamics (PSID; \url{https://psidonline.isr.umich.edu}) is a population based survey study focusing on employment, income, wealth, expenditures, health, marriage, childbearing, child development, philanthropy, and education.
PSID began in 1968 with a U.S. nationally representative sample of 5,000 families and more than 18,000 individuals.
As of mid-2015, PSID had generated more than 3,900 publications. 
The data is available publicly to registered users who agree to the terms of use conditions that ensure participants are not reidentified and that users properly cite and acknowledge use of the data. 
Some restricted data elements (e.g., geospatial data, school-level identifiers) are only available subject to a formal restricted use agreement between the investigator and the data owner, the University of Michigan.

The National Longitudinal Study of Adolescent Health (Ad Health; \url{http://www.nichd.nih.gov/news/releases/Pages/adolescent.aspx}) began as a survey study undertaken in response to a Congressional mandate designed to measure the multiplicity of influences on adolescents' health and unhealthy behaviors.
Parts of the Ad Health cohort have been followed into young adulthood with four in-home interviews, and in a recent wave, biomarker data from a blood test has been made available. 
The Ad Health data are archived at ICSPR and are available for public use.
As of mid-2015, more than 5,000 publications had arisen from the Add Health dataset.

The NICHD Study of Early Child Care and Youth Development (SECCYD; \url{https://www.nichd.nih.gov/research/supported/Pages/seccyd.aspx}) prospectively followed the experiences of 1,364 infants born at 10 locations throughout the United States in 1991.
The goal was to learn more about the kind and quality of child care experiences and the effects of different child care settings on developmental outcomes. 
Measures included video, surveys and interviews, behavioral tests, biomarkers such as salivary cortisol, blood pressure, anthropometrics, and demographic information.
Unlike the Ad Health study, data in SECCYD access are restricted, available only by application, with approval granted on a case-by-case basis.
Staff at the National Institute of Child Health and Human Development manage the data access and approval process.
The SECCYD has generated more than 214 publications as of mid-2015.

Investigator-initiated and managed, the Family Life Project (FLP; \url{http://flp.fpg.unc.edu}) used a population-based approach to study the early development of children living in rural, largely poor communities in North Carolina and Pennsylvania. 
A birth cohort of 800 children were studied in a series of home, child care visits, and phone calls when the infants were 2, 6, 15, 24, and 36 months of age. 
Data included interviews, questionnaires, videotaped interactions, and biological samples yielding information about economic and community factors, family, home, language and cognition, stress, and genotype.
Data from Phase I of the FLP are archived at ICPSR, but are subject to restrictions on access due to the sensitive and identifiable nature of the data.
The FLP site lists about 50 publications through 2013.
